---
title: "Closed Testing Procedure"
author: "Paul Jordan"
date: "`r Sys.Date()`"

output: html_vignette
header_includes:
- \usepackage{amsmath}
- \usepackage{amssymb}

vignette: >
  %\VignetteIndexEntry{Closed Testing Procedure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

references:
- id: Bauer1991
  title: Multiple Testing in Clinical Trials
  author:
  - family: Bauer
    given: Peter
  container-title: Statistics in Medicine
  volume: 10
  issue: 4
  publisher: Nature Publishing Group
  page: 261-263
  type: article-journal
  issued:
    year: 1991

- id: Dmit2010
  title: Multiple Testing Problems in Pharmaceutical Statistics 
  author:
  - family: Dmitrienko
    given: Alex
  - family: Tamhane, A.
  - family: Bretz, F.  
  publisher: Chapman & Hall
  type: book
  issued:
    year: 2010

- id: Bretz2011
  title: Multiple Comparisons Using R
  author:
  - family: Bretz
    given: Frank
  - family: Hothorn, T.
  - family: Westfall, P.
  publisher: CRC Press
  page: 261-263
  type: book
  issued:
    year: 2011
    
- id: Gabriel1976
  title: On Closed Testing Procedures with Special reference to Ordered Analysis of Variance
  author:
  - family: Marcus
    given: Ruth
  - family: Peritz, E.
  - family: Gabriel, K.R.
  container-title: Biometrika
  volume: 63
  page: 655-660
  type: article-journal
  issued:
    year: 1976

- id: emmeans
  title: Estimated Marginal Means, aka Least-Squares Means
  author:
  - family: Russell
  - family: Lenth
  issued:
    year: 2020
---


```{r setup, include = FALSE}
library(knitr);library(CTP)
knitr::opts_chunk$set(collapse = TRUE, out.width="100%"
											 ,fig.width=12,fig.height=8,dev.args = list(pointsize=25) )
```


# The Closure Principle

The closure principle is a way to protect the type-I error from multiple testing. Here, we follow the the description in  [@Bretz2011].
   It consists of four steps:
   
1. Definition of a set ${H} = \{H_1,\ldots, H_n\}$ of elementary hypotheses.

2. Construction of the closure set ("Hypothesis Tree").

$$\overline{H} = \left \{ H_I =\bigcap_{i \in I}H_i :  \quad I \subseteq \{1,\ldots,n\} \right \} $$
$$(\text{all intersection hypotheses} H_I ).$$

3. Construction of a local level-$\alpha$ test for each $H_I \in \overline{H}$.

4. Rejection of $H_i$, if *all* null hypotheses $H_I \in \overline{H}$ with $i \in I$ are rejected at at the local level $\alpha$.


## Adjusted p-values

As the null hypothesis $H_i$ is rejected only if the null hypotheses $H_I \in \overline{H}$ with $i \in I$ are rejected (see point 4. above), the adjusted p-value $p_{adj;i}$ for $H_i$ is defined as:

1. Denote with $p_I$ the p-value for a given intersection hypothesis $H_I, \quad I \subseteq \{1, \ldots,n\}$.
2. Then, $p_{adj;i}=\max_\limits{I:i\in I} p_I,\quad i=1,\ldots , n$. 

# Implementation

The package was designed in partcular for treatment comparisons in ANOVA-like situations.

## Closure set

The hypothesis tree of the closed testing procedure is created using the function `IntersectHypotheses`.

## Local tests for a given "hypothesis tree"
		  
In the case of single hypotheses (i.e. if the hypothesis can be described by a single integer vector e.g. **(1,3,5)** the test (F-Test, Kruskal-Wallis-test, probability test, logrank test, ....) is applied directly.
  
For combined hypotheses (i.e. for hypotheses described by several non-overlapping integer vectors eg. **(1,2), (3,4)**, The procedure differs for (generalised) linear hypotheses and other tests.  
  
In the case of generalised linear hypotheses, the contrast matrices for the single hypotheses included are combined and these contrasts are tested simultaneously. Henceforth, functions from the package `emmeans`[@emmeans] are used, as for all othe linear and generalised linear hypotheses. For all other tests, first the p-values $p_1, p_2, \ldots ,p_m$ for the single hypotheses are calculated, and then these are combined by Fisher's combination rule:

If all $m$ hypotheses are assumed to be independent, the test statistics $X$ follows under $H_0$ a $\chi^2$-distribution with $2m$ degrees of freedom:
$$ X=-2\sum_{i=1}^{m}\ln(p_i) \sim \chi_{2m}^2$$
from which a p-value for the global hypothesis can be easily obtained.

In the case of trend tests, the same type of test is applied for all intermediate single tests. 

## Adjusted p-values
  
Finally the p-values for the elementary *hypotheses* are adjusted by calculating the maximum of the p-values from the hypotheses in the *testing set*  of the respective hypothesis.

The function `AnalyseCTP` calculates all local p-values and the adjusted p-values for all *elementary* hypotheses.

With the function `Adjust_raw`, it is also possible to use p-values that have been calculated by other functions or software to calculate the adjusted p-values.

## Testing set for a specific elementary hypothesis

The testing set for a specific elementary hyothesis can be printed by the function `TestingSet`.

```{r}
library(CTP)
		Pairwise <- IntersectHypotheses(list(c(1,2), c(1,3),
		                                c(1,4), c(2,3), c(2,4), c(3,4)))
    Set24    <- TestingSet(Pairwise,"[24]")
    Set24
```

## Comparing means

The dataframe `pasi` comprises the changes in PASI-score (Psoriasis Area and Severity Index) from baseline
within two month in 72 patients treated with three different doses of  Etretin or Placebo in a double blind study. 
 
The elementary hypotheses 1:2, 1:3, 1:4 are tested simultaneously using the F-Test
i.e. $H_1: \mu_1=\mu_2$, $H_2: \mu_1=\mu_3$ and $H_3: \mu_1=\mu_4$ simultaneously. 
The groups with levels 2,3 and 4 are compared to the control (Placebo) group (level 1). In this specific example, the adjusted and unadjusted p-values are the same. All doses show a significant effect compared to Placebo.

```{r }
library(CTP)

data(pasi)
three.to.first <- IntersectHypotheses(list(1:2,c(1,3),c(1,4)))
Display(three.to.first,Type="s",arrow=TRUE)
pasi.ctp.F1    <- AnalyseCTP(three.to.first,pasi.ch~dose,pasi)

summary(pasi.ctp.F1)
Display(pasi.ctp.F1)
```


## Another hypothesis structure

Testing the elementary hypotheses 1:2, 2:3, 3:4 simultaneously using the F-Test, i.e. testing $H_1: \mu_1=\mu_2$, $H_2: \mu_2=\mu_3$ and $H_3: \mu_3=\mu_4$ simultaneously. This provides quite different results (compared to `pasi.ctp.F1`): No further improvement for higher doses.

```{r}
dose.steps4 <- IntersectHypotheses(list(1:2,2:3,3:4))
Display(dose.steps4,arr=TRUE)



pasi.ctp.F2 <- AnalyseCTP(dose.steps4,pasi.ch~dose,pasi)
summary(pasi.ctp.F2)
Display(pasi.ctp.F2)
```

## Other tests

For the same hypothesis structure, other tests can also be used:

### Generalized Linear Models

As an example, a positive response is defined as a change from baseline PASI score greater than 50. The new variable `Resp` has then the value 1 if *pasi.ch > 50* or 0 otherwise. The corresponding model is chosen to be a generalised linear model with logit-link, as implemented in `glm`.  

```{r}
pasi$Resp <- ifelse(pasi$pasi.ch > 50,1,0)

pasi.ctp_bin <-AnalyseCTP(three.to.first,Resp~dose,pasi,test.name="glm",family="binomial")
summary(pasi.ctp_bin)
```

### Kruskal-Wallis test of trend for all single hypotheses

```{r}
pasi.ctp.K <- AnalyseCTP(dose.steps4,pasi.ch~dose,pasi, test="kruskal")
summary(pasi.ctp.K)
Display(pasi.ctp.K)
```

### Jonckheere-Terpstra test of trend for all single hypotheses

```{r,warning=FALSE,message=FALSE}
pasi.ctp.J1 <- AnalyseCTP(dose.steps4,pasi.ch~dose,pasi, test="jonckheere",alternative="increasing")
summary(pasi.ctp.J1)
```

## Proportions

The data set `colorectal` contains the response rates from a dose finding study in metastatic colorectal cancer. Two doses of the experimental drug were compared to the standard treatment. The response rates in the two dose groups are compared to the control responder rate using both, the $\chi^2$-test and Fisher's exact test.

```{r}
two.to.first<- IntersectHypotheses(list(1:2,c(1,3)))

Display(two.to.first,Type="s",main="two vs control",arrow=TRUE)

#The two elementary hypotheses  are tested after comparing the three proportions globally.

data(colorectal)
colorectal.ctp <-AnalyseCTP(two.to.first,responder~dose,data=colorectal, test="prob")
summary(colorectal.ctp)
Display(colorectal.ctp,Type="t")

colorectal.chisq <-AnalyseCTP(two.to.first,responder~dose,data=colorectal, test="chisq")
summary(colorectal.chisq, digits=1)
```

## Survival Analysis with the logrank test

 This example uses the sample dataset `ovarian` from the package `survival`. The overall survival curves of the two treatments `rx` do not differ significantly:
 
 ```{r}
 library(survival)
		data(ovarian)
		
		print(survdiff(Surv(futime,fustat)~rx, data=ovarian))
```

Together with the performance subgroups `ecog=1` and `ecog=2` , a factor "subgroups" defined by the combinations of the performance measure `ecog.ps` and the treatment `rx`.

```{r}		
ovarian$subgroups <- as.factor(10*ovarian$ecog.ps+ovarian$rx)
print(head(ovarian))
```

Then, the treatment differences within the performance subgroups ecog=1 and ecog=2 are compared. I.e. the elementary hypotheses are *subgroup11=subgroup12* and *subgroup21=subgroup22* or $\{(1,2),(3,4)\}$.

```{r}		
comb.sub  <- IntersectHypotheses(list(c(1,2),c(3,4)))
#Display(comb.sub)
ovar.ctp  <-AnalyseCTP(comb.sub,Surv(futime,fustat)~subgroups, ovarian, test="lgrank")
summary(ovar.ctp)
Display(ovar.ctp)
```		

## Comparing means when a covariate is included

In a study with diabetes type II patients (dataset `glucose`), three doses of a drug are compared to a placebo. The primary variable is the change of fasting plasma glucose from baseline. Fasting plasma glucose at baseline is included into the model as covariate (only implemented for linear and generalised linear models).

```{r}		
data(glucose)
glucose.ctp <- AnalyseCTP(three.to.first,GLUCOSE.CHANGE~GLUCOSE.BLA+DOSE,
                         data=glucose, factor.name="DOSE")
summary(glucose.ctp)
Display(glucose.ctp,Type="s")
```

## Large hypothesis trees

Whith an increasing number of hypotheses to test, the graphical display may become quite confusing:


```{r}
G <- factor(rep(1:5,each=4)	)			
y <- rnorm(20)
Y <- data.frame(G,y)
				
xxx <- IntersectHypotheses(list(1:2,c(1,3),c(1,4),c(1,5),c(2,5),c(3,4)))
summary(xxx)
Display(xxx)
```

## "External" p-values

It is possible to:

* Use `IntersectHypotheses` to create the closure set and `Display` to plot the corresponding hypothesis tree.
* Calculate the p-values for the local hypotheses with other functions or software.
* Import these p-values.
* Use `Adjust_raw` to calculte the adjusted p-values for the elementary hypotheses. `Adjust_raw` delivers an object of class `ctp`.
* Use `summary` and `Display` on this object.

```{r}
		Pairwise <- IntersectHypotheses(list(c(1,2), c(1,3), c(1,4), c(2,3), c(2,4), c(3,4)))
		Display(Pairwise)
		summary(Pairwise)
		
		# the vector of p-values calculated by another software
		# (Exampe from Prof. John M. Lachin, The Biostatistics Center Rockville MD)
		
		p.val <- c(
		  0.4374,
		  0.6485,
		  0.4103,
		  0.2203,
		  0.1302,
		  0.6725,
		  0.4704,
		  0.3173,
		  0.6762,
		  0.7112,
		  0.2866,
		  0.3362,
		  0.2871,
		  0.4633)

		result <- Adjust_raw(Pairwise, p.value=p.val)
		
		summary(result,digits=3)	
				
# details may be documented
		
		result <- Adjust_raw(Pairwise, p.value=p.val
		                     ,dataset.name="my Data", factor.name="Factor"
		                     ,factor.levels=c("A","B","C","D"), model=y~Factor
		                     ,test.name="my Test")
		
		summary(result,digits=3)
		Display(result) 
```

---
nocite: | 
  @Gabriel1976, @Bauer1991, @Dmit2010
...

## References
